Waggle dance distributions quantify collective foraging in honey bee colonies: Supplementary material

### 2022-04-11 analysis/manuscript/paper.Rmd
updated to use 2017 data, updated ms and si and methods. Lots of stuff as we approach final edits for submission

For 17 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging (Fig. 3a). In all but one of these cases, the collective model had a good fit (using a Kolmogorov-Smirnov statistic, see methods) to the empirical waggle run durations, whereas for 9 of these sites, the distribution predicted by the scout model was significantly different to the observed data (Fig. 3b). The sites STU and ZSL shown in Figs 3c-d are representative examples showing the model fits where the collective (ZSL) and individual (STU) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

In all twenty sites the collective model provided a higher maximum likelihood estimate (MLE) than the individual model, however, this is anticipated as the individual model is a subset of the collective model. In 17 of the sites the collective model accounted for over 95% of Akaike weights compared to the individual models fit (supplementary table 1). In 3 of the sites the individual model explained a higher proportion of Akaike weight than the collective model (supplementary table 1) and so honeybee foraging behaviour in these colonies is better explained by the individual rather than collective foraging model.

```{r, fig.cap="Figure 4. **The honey bee foraging model fitted to data from 20 hives**.  (a) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 17 hives for the collective foraging model provided best explanation (black circles) or for 3 hives the individual search model provided the best explanation(red triangles) as indicated by lowest AIC score. (b). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. (c) For the STU hive the individual model (blue line) provided a better fit than the collective foraging model (red line). (d). For the ZSL hive the collective foraging model (red line) outperformed the individual model (blue line). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../results/figures/manuscript_figures/fig4.png")
```

Our results show that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], although it has also been shown that bees may cease dance following if it is proving unproductive [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individual's foraging behaviour over a landscape. As our results show that individual foraging accurately describes colony foraging in three different sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

Further quantification of the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of scout trips, $p$, for each site. Since our sites varied in land-use characteristics and potentially thus forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see methods) to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances within agri-rural and urban environments. As we have a small sample size (10 urban and 10 agri-rural sites), we used jackknife resampling to evaluate the robustness of our results to influential points (see methods, supplementary material)

In the agri-rural environments the first principle component is a combination of land use types that explained ~52% of the variation in the proportion of scouts (beta regression: $R^2 = 0.52$, $\phi$ = 3.7, p < 0.05, Fig. 4. A). This principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage); note that it also correlates negatively with non-agricultural unmanaged green space and water, but together these represent less than 3% of land-use (Fig. 5b). These land-use types maintain a significant correlation with the first principle component over the jackknifed PLS (Fig. 4. B) indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases. Arable land in the UK is typically considered nutritionally poor for bees (note that oilseed rape fields were not included within this category), while there is evidence to suggest that the residential areas that were captured within the “built-up” category are forage-rich hotspots, typically supporting relatively high bee diversity and abundance within gardens [@Baldock2019].

@@

In in the urban environment, the first principal component explained ~59% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.59, $\phi$ = 6.2, p < 0.05, Fig. 5c). Similarly to the agri-rural environments, this component correlates negatively with a significant residential category (sparse residential lands; 35% of land cover; Table 1), such that scout trips would appear to increase as forage availability decreases. However, we have limited confidence in this pattern because (a) this component correlates positively with parks, allotments and cemeteries (8% of land cover), which are typically also forage-rich [@Baldock2019] (b) the jackknifed PLS identified it being influenced strongly by a single site, the removal of which causes significant distribution to both of these relationships (Supplementary Material).

>>>>

For 16 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging ($wAIC > 0.95$, Fig. 4a). In the other 4 sites, despite the collective model having the higher maximum likelihood estimate, the individual model had higher Akaike weights and so is judged to be more parsimonious (Supplementary Table 1). In all but one site, the collective model had a good fit (using a Kolmogorov-Smirnov statistic of $p > 0.05$, see methods) to the empirical waggle run durations (Fig. 4b), whereas the scout model was significantly different to the observed data in 8 sites (Kolmogorov-Smirnov statistic $p < 0.05$, Fig. 4b). The sites STU and ZSL shown in Figs 4c-d are representative examples showing the model fits where the collective (ZSL) and individual (STU) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

```{r, fig.cap="Figure 4. **The honey bee foraging model fitted to data from 20 hives**.  (a) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 16 hives for the collective foraging model provided best explanation (black circles) or for 4 hives the individual search model provided the best explanation(red triangles) as indicated by lowest AIC score. (b). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. (c) For the STU hive the individual model (blue line) provided a better fit than the collective foraging model (red line). (d). For the ZSL hive the collective foraging model (red line) outperformed the individual model (blue line). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../results/figures/manuscript_figures/fig4.png")
```

Our results show that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], although it has also been shown that bees may cease dance following if it is proving unproductive [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individual's foraging behaviour over a landscape. As our results show that individual foraging accurately describes colony foraging in 11 different sites, and more parsimoniously than a model of collective foraging in 4 sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

Further quantification of the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of scout trips, $p$, for each site. Since our sites varied in land-use characteristics and potentially thus forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see methods) to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances within agri-rural and urban environments. As one of our sites (ROT) was not well fit by either of our models, this sites was removed from the PLS analysis. Due to our small sample size (10 urban and 9 agri-rural sites), we used jackknife resampling to evaluate the robustness of our results to influential points (see methods, supplementary material).

In the agri-rural environments the first principle component is a combination of land use types that explained ~73% of the variation in the proportion of scouts (beta regression: $R^2 = 0.73$, $\phi$ = 4.9, p < 0.05, Fig. 5a). This principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage); note that it also correlates negatively with non-agricultural unmanaged green space and water, but together these represent less than 3% of land-use (Table 1; Fig. 5b). These land-use types maintain a significant correlation with the first principle component over the jackknifed PLS (Fig. 5b), with the exception of non-agricultural unmanaged green space which sits on the boarder, indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases. Arable land in the UK is typically considered nutritionally poor for bees (note that oilseed rape fields were not included within this category), while there is evidence to suggest that the residential areas that were captured within the “built-up” category are forage-rich hotspots, typically supporting relatively high bee diversity and abundance within gardens [@Baldock2019].

@@

In in the urban environment, the first principal component explained ~73% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5c). This component correlates positively with parks, allotments and cemeteries (8% of land cover; Table 1) and negatively with railways, however this accounts for ~1% of land cover (Table 1). The dominant land by coverage , sparse residential (land coverage ~35%; Table 1), has a significant negative correlation with the first principle component for some sites removed over the jackknife sampling (Supplementary Fig. 2.), however, it is overall not a significant contributor to the variance in the loadings (Fig. 5c.). These results seem to be in contrast to what was observed in the urban landscapes as parks, allotments and cemeteries are typically forage-rich for honeybees [@Baldock2019]. However, we have limited confidence in these results as (a) parks, allotments and cemeteries account for just 8% of land cover and (b) the jackknifed PLS identified substantial variation in the loadings identified for the first principle components (Supplementary Fig. 3.).

### 2022-04-11 analysis/manuscript/supplementary_material.Rmd
updated to use 2017 data, updated ms and si and methods. Lots of stuff as we approach final edits for submission

```{r, all-model-fits, fig.width=10,fig.height=11, fig.cap="Supplementary Fig 1. log Compliment cumulative frequency distribution of waggle dance durations and fits of the collective (red line) and individual (blue line) models for all 20 sites."}
plots <- map(all_sites, 2)
plots <- plots %>%
  map(
    ~ {
      .x + theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank()
      )
    }
  )

@@

@@

```{r, rural-pls-loadings, fig.width=10,fig.height=11, fig.cap="Loadings of PLS calculated for each site removed for the agri-rural sites. Each plot shows the loadings of the first principle component with that site removed from the analysis, showing the individual points making up the overall box plot loadings in Fig 4b."}
knitr::include_graphics("../results/figures/SI_figures/agrirural_jk_individual_loadings.png")
```

```{r, urban-pls-loadings, fig.width=10,fig.height=11, fig.cap="Loadings of PLS calculated for each site removed for the urban sites. Each plot shows the loadings of the first principle component with that site removed from the analysis, showing the individual points making up the overall box plot loadings in Fig 4d."}
knitr::include_graphics("../results/figures/SI_figures/urban_jk_individual_loadings.png")
```

>>>>

```{r, all-model-fits, fig.width=10,fig.height=11, fig.cap="Supplementary Fig 1. **Cumulative frequency distribution** of waggle dance durations and fits of the collective (red line) and individual (blue line) models for all 20 sites."}
plots <- map(all_sites, 2)
plots <- plots %>%
  map(
    ~ {
      .x + theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank()
      )
    }
  )

@@

write.csv(df2, "../results/modelfits_allsites.csv")

@@

```{r, rural-pls-loadings, fig.width=10,fig.height=11, fig.cap="Supplementary Fig 2. **Loadings of PLS calculated for each site removed for the agri-rural sites**. Each plot shows the loadings of the first principle component with that site removed from the analysis, showing the individual points making up the overall box plot loadings in Fig 4b."}
knitr::include_graphics("../results/figures/SI_figures/agrirural_jk_individual_loadings.png")
```

```{r, urban-pls-loadings, fig.width=10,fig.height=11, fig.cap="Supplementary Fig 3. **Loadings of PLS calculated for each site removed for the urban sites**. Each plot shows the loadings of the first principle component with that site removed from the analysis, showing the individual points making up the overall box plot loadings in Fig 4d."}
knitr::include_graphics("../results/figures/SI_figures/urban_jk_individual_loadings.png")
```
